Random Generation of Rich \\ Abstract Data Type Values

### 2019-02-01 paper/casestudies.tex
Spell checking

We evaluate how including additional structural information when generating a
set of random test cases (often referred as a \emph{corpus}) affects the code
coverage obtained when testing a given target program.
%
For that, we considered two external programs which expect highly structured
inputs, namely \emph{GNU CLISP}---the GNU Common Lisp
compiler\footnote{\href{https://www.gnu.org/software/gcl/}{https://www.gnu.org/software/gcl/}},
and \emph{HTML Tidy}
\footnote{\href{http://www.html-tidy.org}{http://www.html-tidy.org}}---a well
known HTML refactoring and correction utility.
%
We remark that these applications are not written in Haskell.
%
However, there exist Haskell libraries defining ADTs encoding their input
structure, i.e., Lisp and HTML values respectively. These libraries are:
\emph{hs-zuramaru}\footnote{\href{http://hackage.haskell.org/package/zuramaru}{http://hackage.haskell.org/package/zuramaru}},
implementing an embedded Lisp intepreter for a small subset of this programming
language, and
\emph{html}\footnote{\href{http://hackage.haskell.org/package/html}{http://hackage.haskell.org/package/html}},
defining a combinator library for constructing HTML values.
%
These libraries also come with serialization functions to map Haskell values
into corresponding test case files.

@@

In this case study we included the structural information present on the abstract
interface of this library into the generation specification of random HTML
values, resulting in a generation process that randomly picked among 4 data
constructors and 163 abstract functions.
%
With this large amount of additional structural information, we observed an
increase of up to $83\%$ in the code coverage obtained with \dragenp with
respect to the one observed with \dragen (see Figure \ref{fig:coverage} (b)).
%
A manual inspection of the corpora generated with each tool revealed us that, in
general terms, the test cases generated with \dragen rarely represent
syntactically correct HTML values, consisting to a large extent of random
strings within and between HTML tag delimeters (|"<"|, |">"| and |"/>"|).
%
On the other hand, test cases generated with \dragenp encode much more
interesting structural information, being mostly syntactically correct.
%
We found that, in many cases, the test cases generated with \dragenp were
parsed, analysed and reported as valid HTML values by the target application.

>>>>

We evaluate how including additional structural information when generating a
set of random test cases (often referred as a \emph{corpus}) affects the code
coverage obtained when testing a given target program.
%
For that, we considered two external programs which expect highly structured
inputs, namely \emph{GNU CLISP}---the GNU Common Lisp
compiler\footnote{\href{https://www.gnu.org/software/gcl/}{https://www.gnu.org/software/gcl/}},
and \emph{HTML Tidy}
\footnote{\href{http://www.html-tidy.org}{http://www.html-tidy.org}}---a well
known HTML refactoring and correction utility.
%
We remark that these applications are not written in Haskell.
%
However, there exist Haskell libraries defining ADTs encoding their input
structure, i.e., Lisp and HTML values respectively. These libraries are:
\emph{hs-zuramaru}\footnote{\href{http://hackage.haskell.org/package/zuramaru}{http://hackage.haskell.org/package/zuramaru}},
implementing an embedded Lisp interpreter for a small subset of this programming
language, and
\emph{html}\footnote{\href{http://hackage.haskell.org/package/html}{http://hackage.haskell.org/package/html}},
defining a combinator library for constructing HTML values.
%
These libraries also come with serialization functions to map Haskell values
into corresponding test case files.

@@

In this case study we included the structural information present on the abstract
interface of this library into the generation specification of random HTML
values, resulting in a generation process that randomly picked among 4 data
constructors and 163 abstract functions.
%
With this large amount of additional structural information, we observed an
increase of up to $83\%$ in the code coverage obtained with \dragenp with
respect to the one observed with \dragen (see Figure \ref{fig:coverage} (b)).
%
A manual inspection of the corpora generated with each tool revealed us that, in
general terms, the test cases generated with \dragen rarely represent
syntactically correct HTML values, consisting to a large extent of random
strings within and between HTML tag delimiters (|"<"|, |">"| and |"/>"|).
%
On the other hand, test cases generated with \dragenp encode much more
interesting structural information, being mostly syntactically correct.
%
We found that, in many cases, the test cases generated with \dragenp were
parsed, analyzed and reported as valid HTML values by the target application.

### 2019-02-01 paper/finalremarks.tex
Spell checking

In the future we aim to extract stronger invariants from the user codebase into
our automatically derived generators, as well as evaluating how introducing
random mutations into structurally complex values might improve testing
performace, specially when generating ``almost correct'' random data.

>>>>

In the future we aim to extract stronger invariants from the user codebase into
our automatically derived generators, as well as evaluating how introducing
random mutations into structurally complex values might improve testing
performance, specially when generating ``almost correct'' random data.

### 2019-02-01 paper/randomtesting.tex
Spell checking

\end{code}
%
The type |Html| allows to build pages via four possible constructions:
|Text|---which represents plain text values---, |Single| and |Tag|---which
represent singular and paired Html tags, respectively---, and |Join|---which
concats two Htmls pages one after another.
%
% In the previous definition, we declare |Html| as the sum of four possible
% constructions: |Text| represents plain text values. |Sing| and |Tag| represent
% singular and paired Html tags, and |Join| concats two expression one after
% another.
% %
% We only encode a very small subset of the actual Html specification for
% illustrative reasons.
% %
In Haskell, |Text|, |Single|, |Tag|, and |Join| are known as \emph{data
  constructors} (or constructors for short) and are used to distiguish which
variant of the ADT we are constructing.
%
Each data constructor is defined as a product of zero or more types known as
\emph{fields}.
%
For instance, |Text| has a field of type |String|, whereas |Join| has two
recursive fields of type |Html|.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this representation, the example page \texttt{<html>hello<hr>bye</html>}
can be encoded as:
%
\begin{code}
Tag "html" (Join (Join
  (Text "hello") (Single "hr")) (Text "bye"))
\end{code}
%
% Additionally, we can define a function |render| to serialize |Html| values as
% follows:
% %
% \begin{code}
% render :: Html -> String
% render (Text t) = t
% render (Sing t) = "<" ++ t ++ ">"
% render (Join x y) = render x ++ render y
% render (Tag t x)
%   = "<" ++ t ++ ">" ++ render x ++ "</" ++ t ++ ">"
% \end{code}

@@

In order to generate random ADTs values, most approaches require users to provide
a random data generator for each ADT definition.
%
This is a cumbersome and error prone task that usually \emph{follows closely
  ADTs structures}.
%
For instance, consider the following definition of a \quickcheck random
generator for the type |Html|:
%
\begin{code}
genHtml = sized (\size ->
  if size == 0
  then frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString) ]
  else frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString)
  ,  (4,  Tag     <$> genString <*> smaller genHtml)
  ,  (3,  Join    <$> smaller genHtml <*> smaller genHtml) ])
\end{code} %$
%
We use the Haskell syntax |[]| and |(,)| for denoting lists and pairs of
elements, respecitvely (e.g., |[(1,2),(3,4)]| is a list of pairs of numbers.)
%
The random generator |genHtml| is defined using \quickcheck's function |sized|
to paremetrize the generation process up to an external natural number known as
the \emph{generation size}---captured in the code with variable |size|.
%
This parameter is chosen by the user, and it is used to limit the maximum amount
of recursive calls that this random generator can perform and thus ensuring
the termination of the generation process.
%
When called with a positive generation size, this generator can pick to generate
among any |Html| data constructor with an explicitly \emph{generation frequency}
that can be chosen by the user---in this example, 2, 1, 4 and 3 for |Text|,
|Single|, |Tag|, and |Join|, respectively.
%
When it picks to generate a |Text| or a |Single| data constructor, it also
generates a random |String| value using the standard QuickCheck generator
|genString|.% (\quickcheck provides standard random generators for most base types
% like |String|, |Int|, |Bool|, etc.)
\footnote{%
  The operators |<$>| and |<*>| are used in Haskell to combine values obtained
  from calling random generators and they are not particularly relevant for the
  point being made in this work.}
%$
On the other hand, when it picks to generate a |Join| constructor, it also
generates two independent random sub-expression recursively, decreasing the
generation size by a unit on each recursive invocation (|smaller genHtml|).
%
The case of random generation of |Tag| constructors follows analogously.
%
This random process keeps calling itself recursively until the generation size
reaches zero, where the generator is constrained to pick among terminal data
constructors, being |Text| and |Single| the only possible choices in our
particular case.

>>>>

\end{code}
%
The type |Html| allows to build pages via four possible constructions:
|Text|---which represents plain text values---, |Single| and |Tag|---which
represent singular and paired HTML tags, respectively---, and |Join|---which
concatenates two HTMLs pages one after another.
%
% In the previous definition, we declare |Html| as the sum of four possible
% constructions: |Text| represents plain text values. |Sing| and |Tag| represent
% singular and paired Html tags, and |Join| concats two expression one after
% another.
% %
% We only encode a very small subset of the actual Html specification for
% illustrative reasons.
% %
In Haskell, |Text|, |Single|, |Tag|, and |Join| are known as \emph{data
  constructors} (or constructors for short) and are used to distinguish which
variant of the ADT we are constructing.
%
Each data constructor is defined as a product of zero or more types known as
\emph{fields}.
%
For instance, |Text| has a field of type |String|, whereas |Join| has two
recursive fields of type |Html|.
%
In general, we will say that a data constructor with no recursive fields is
\emph{terminal}, and \emph{non-terminal} or \emph{recursive} if it has at least
one field of such nature.
%
With this representation, the example page \texttt{<html>hello<hr>bye</html>}
can be encoded as:
%
\begin{code}
Tag "html" (Join (Join
  (Text "hello") (Single "hr")) (Text "bye"))
\end{code}
%
% Additionally, we can define a function |render| to serialize |Html| values as
% follows:
% %
% \begin{code}
% render :: Html -> String
% render (Text t) = t
% render (Sing t) = "<" ++ t ++ ">"
% render (Join x y) = render x ++ render y
% render (Tag t x)
%   = "<" ++ t ++ ">" ++ render x ++ "</" ++ t ++ ">"
% \end{code}

@@

In order to generate random ADTs values, most approaches require users to provide
a random data generator for each ADT definition.
%
This is a cumbersome and error prone task that usually \emph{follows closely the
  structure of the ADTs}.
%
For instance, consider the following definition of a \quickcheck random
generator for the type |Html|:
%
\begin{code}
genHtml = sized (\size ->
  if size == 0
  then frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString) ]
  else frequency
  [  (2,  Text    <$> genString)
  ,  (1,  Single  <$> genString)
  ,  (4,  Tag     <$> genString <*> smaller genHtml)
  ,  (3,  Join    <$> smaller genHtml <*> smaller genHtml) ])
\end{code} %$
%
We use the Haskell syntax |[]| and |(,)| for denoting lists and pairs of
elements, respectively (e.g., |[(1,2),(3,4)]| is a list of pairs of numbers.)
%
The random generator |genHtml| is defined using \quickcheck's function |sized|
to parameterize the generation process up to an external natural number known as
the \emph{generation size}---captured in the code with variable |size|.
%
This parameter is chosen by the user, and it is used to limit the maximum amount
of recursive calls that this random generator can perform and thus ensuring
the termination of the generation process.
%
When called with a positive generation size, this generator can pick to generate
among any |Html| data constructor with an explicitly \emph{generation frequency}
that can be chosen by the user---in this example, 2, 1, 4 and 3 for |Text|,
|Single|, |Tag|, and |Join|, respectively.
%
When it picks to generate a |Text| or a |Single| data constructor, it also
generates a random |String| value using the standard QuickCheck generator
|genString|.% (\quickcheck provides standard random generators for most base types
% like |String|, |Int|, |Bool|, etc.)
\footnote{%
  The operators |<$>| and |<*>| are used in Haskell to combine values obtained
  from calling random generators and they are not particularly relevant for the
  point being made in this work.}
%$
On the other hand, when it picks to generate a |Join| constructor, it also
generates two independent random sub-expression recursively, decreasing the
generation size by a unit on each recursive invocation (|smaller genHtml|).
%
The case of random generation of |Tag| constructors follows analogously.
%
This random process keeps calling itself recursively until the generation size
reaches zero, where the generator is constrained to pick among terminal data
constructors, being |Text| and |Single| the only possible choices in our
particular case.

### 2019-02-01 paper/relatedwork.tex
Spell checking

%
% Boltzmann models
%
Boltzmann models \cite{Duchon2004} are a general approach to randomly generating
combinatorial structures such as trees and graphs, closed simply-typed lambda
terms, etc.
%
A random generator built around such models uniformly generates values of a
target size with a certain size tolerance.
%
However, it has been argued that this approach has theoretical and practical
limitations in the context of software testing \cite{feldt2013}.
%
In a recent work, Bendkowski provides a framework called \emph{boltzmann-brain}
to specify and synthesize standalone Haskell random generators based on
Boltzmann models \cite{bendkowski2018}.
%
This framework mixes parameter tuning and rejection of unwanted samples to
approximate the desired distribution of values according to user demands.
%
In this approach, the overall discard ratio depends on how constrained the
desired distribution is.
%
On the other hand, our work is focused on approximating the desired distribution
as much as possible via parameter optimization, without discarding any generated
value at runtime.
%
Altough promising, we found several difficulties to compare both approaches in
practice due that \emph{boltzmann-brain} is considered a standalone utility that
produces self-contained samplers.
%
In this light, data specifications have to be manually written using a special
syntax, and cannot include Haskell ground types like |String| or |Int|,
difficulting the integration of this tool to existing Haskell codebases like the
ones we consider in this work.

%
% GodelTest
%
From the practical point of view, Feldt and Poulding propose \emph{G\"odelTest}
\cite{feldt2013}, a search-based framework for generating biased data.
%
% \emph{G\"odelTest} uses non-determinism to generate a wide range of data structures.
%
Similar to our approach, \emph{G\"odelTest} works by optimizing the parameters
governing the desired biases on the generated data.
%
However, the optimization mechanism uses metaheuristic search to find the best
parameters at runtime.
%
\dragenp on the other hand implements an analytic and composable prediction
mechanism that is used only used at compile time to optimize the generation
parameters, thus avoiding perfomring any kind of runtime reinforcement.

%
% DART
%
Directed Automated Random Testing (DART) is a technique that combines random
testing with symbolic execution for C programs \cite{godefroid2005dart}.
%
It requires instrumenting the target programs in order to introduce testing
assertions and obtain feedback from previous testing executions, which is used
to explore new paths in the source code.
%
This technique has been shown to be remarkably useful, altough it forces a
strong coupling between the testing suite and the target code.
%
Our tool intends to provide better random generation of values following an
undirected fashion, without having to instrument the target code, but still
extracting useful structural information from it.

>>>>

%
% Boltzmann models
%
Boltzmann models \cite{Duchon2004} are a general approach to randomly generating
combinatorial structures such as trees and graphs, closed simply-typed lambda
terms, etc.
%
A random generator built around such models uniformly generates values of a
target size with a certain size tolerance.
%
However, it has been argued that this approach has theoretical and practical
limitations in the context of software testing \cite{feldt2013}.
%
In a recent work, Bendkowski provides a framework called \emph{boltzmann-brain}
to specify and synthesize standalone Haskell random generators based on
Boltzmann models \cite{bendkowski2018}.
%
This framework mixes parameter tuning and rejection of unwanted samples to
approximate the desired distribution of values according to user demands.
%
In this approach, the overall discard ratio depends on how constrained the
desired distribution is.
%
On the other hand, our work is focused on approximating the desired distribution
as much as possible via parameter optimization, without discarding any generated
value at runtime.
%
Although promising, we found several difficulties to compare both approaches in
practice due that \emph{boltzmann-brain} is considered a standalone utility that
produces self-contained samplers.
%
In this light, data specifications have to be manually written using a special
syntax, and cannot include Haskell ground types like |String| or |Int|,
difficulting the integration of this tool to existing Haskell codebases like the
ones we consider in this work.

%
% GodelTest
%
From the practical point of view, Feldt and Poulding propose \emph{G\"odelTest}
\cite{feldt2013}, a search-based framework for generating biased data.
%
% \emph{G\"odelTest} uses non-determinism to generate a wide range of data structures.
%
Similar to our approach, \emph{G\"odelTest} works by optimizing the parameters
governing the desired biases on the generated data.
%
However, the optimization mechanism uses meta-heuristic search to find the best
parameters at runtime.
%
\dragenp on the other hand implements an analytic and composable prediction
mechanism that is used only used at compile time to optimize the generation
parameters, thus avoiding performing any kind of runtime reinforcement.

%
% DART
%
Directed Automated Random Testing (DART) is a technique that combines random
testing with symbolic execution for C programs \cite{godefroid2005dart}.
%
It requires instrumenting the target programs in order to introduce testing
assertions and obtain feedback from previous testing executions, which is used
to explore new paths in the source code.
%
This technique has been shown to be remarkably useful, Although it forces a
strong coupling between the testing suite and the target code.
%
Our tool intends to provide better random generation of values following an
undirected fashion, without having to instrument the target code, but still
extracting useful structural information from it.

### 2019-02-01 paper/synthesis.tex
Spell checking

We characterize the numbers of constructors that a random generator produces in
the $n$-th generaton as a vector $G_n$, a vector that groups the number of
constructors of each kind produced in that generation---in our |Html| example,
this vector has four components, i.e., one for each constructor.
%
%We note $G_n$ to the vector of real numbers that represents the distribution of
%
From branching processes theory, the following equation captures the expected
distribution of constructors at the generation $n$, noted $E[G_n]$, as follows:
%
\begin{align}
  E[G_n]^T = E[G_0]^T \cdot M^n
  \label{eqn:prediction}
\end{align}

Vector $E[G_0]$ represents the initial distribuition of constructors that our
generator produces, which simply consists of the generation probability of each
one.
%
The interesting aspect of the prediction mechanism is encoded in the matrix $M$,
known as a the \emph{mean matrix} of this stochastic process.
%
$M$ is a squared matrix with as rows and columns as different data constructors
involved in the generation process.
%
Each element $M_{i,j}$ of this matrix encodes the average number of data
constructors of kind $j$ that gets generated in a given generation, provided that we
generated a constructor of kind $i$ at the previous one.
%
In this sense, this matrix encodes the ``branching'' behavior of our random
generation from one generation to the next one.
%
Each element of the matrix can be automatically calculated by exploiting ADT
definitions. % \cite{DBLP:conf/haskell/MistaRH18}
%
% ---specially the recursive fields---
%
as well as the individual probability of generating each constructor.
%
For instance, the average number of |Text| data constructors that we will
generate provided that we generated a |Join| data constructor on the previous
level is calculated as follows:
%
$$ M_{\mathrm{Join, Text}} = 2\cdot p_{\mathrm{Text}} $$
%
where $2$ is the number of holes present when generating a partial ADT value
|Join| (i.e., |Join hole hole|) and $p_{\mathrm{Text}}$ is the probability of
individually generating the constructor |Text|.
%
This pattern can be used to build the rest of the mean matrix as well.

@@

As another contribution, we found that the whole prediction process can be
factored in terms of two vectors $\beta$ and $\mathcal{P}$, such that $\beta$
represents the number of holes in each partial ADT value that we generate,
whereas $\mathcal{P}$ simply represents the probability of generating that
partial ADT value.% (this probability can be easily calculated based on the
% generation frequencies used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewriten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P}^T)^{n}
\end{align*}
%
% The vector $\beta$ can be obtained by analyzing the shape (namely holes) of each
% source of structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as shown in Fig. \ref{fig:vectors}.
%
We note |simplify1| and |simplify2| to the patterns occurring in the first and
second clauses of |simplify|, respectively.
%(Dashed lines indicate the composition of the diffent sub-vectors
%that are calculated separately).

@@

Note that by varying the shape of the vector $\mathcal{P}$ we can tune the
distribution of our random genrator in a way that can be always characterized
and predicted.
%
\dragenp follows a similar approach as \dragen and uses an heuristic to tune the
generation probabilities of each source of structural information.
%
This is done by runing a simulation-based optimization process at compile-time.
%
This process is parametrized by the desired distribution of values set by the
user.
%
In this manner, developers can specify, for instance, a uniform distribution of
data constructors, pattern matching values and function calls or, alternatively,
a distribution of values with some constructions appearing in a different
proportion as others, e.g., two times more functions calls to |div| than |Join|
constructors.

>>>>

We characterize the numbers of constructors that a random generator produces in
the $n$-th generation as a vector $G_n$, a vector that groups the number of
constructors of each kind produced in that generation---in our |Html| example,
this vector has four components, i.e., one for each constructor.
%
%We note $G_n$ to the vector of real numbers that represents the distribution of
%
From branching processes theory, the following equation captures the expected
distribution of constructors at the generation $n$, noted $E[G_n]$, as follows:
%
\begin{align}
  E[G_n]^T = E[G_0]^T \cdot M^n
  \label{eqn:prediction}
\end{align}

Vector $E[G_0]$ represents the initial distribution of constructors that our
generator produces, which simply consists of the generation probability of each
one.
%
The interesting aspect of the prediction mechanism is encoded in the matrix $M$,
known as a the \emph{mean matrix} of this stochastic process.
%
$M$ is a squared matrix with as rows and columns as different data constructors
involved in the generation process.
%
Each element $M_{i,j}$ of this matrix encodes the average number of data
constructors of kind $j$ that gets generated in a given generation, provided that we
generated a constructor of kind $i$ at the previous one.
%
In this sense, this matrix encodes the ``branching'' behavior of our random
generation from one generation to the next one.
%
Each element of the matrix can be automatically calculated by exploiting ADT
definitions. % \cite{DBLP:conf/haskell/MistaRH18}
%
% ---specially the recursive fields---
%
as well as the individual probability of generating each constructor.
%
For instance, the average number of |Text| data constructors that we will
generate provided that we generated a |Join| data constructor on the previous
level is calculated as follows:
%
$$ M_{\mathrm{Join, Text}} = 2\cdot p_{\mathrm{Text}} $$
%
where $2$ is the number of holes present when generating a partial ADT value
|Join| (i.e., |Join hole hole|) and $p_{\mathrm{Text}}$ is the probability of
individually generating the constructor |Text|.
%
This pattern can be used to build the rest of the mean matrix as well.

@@

As another contribution, we found that the whole prediction process can be
factored in terms of two vectors $\beta$ and $\mathcal{P}$, such that $\beta$
represents the number of holes in each partial ADT value that we generate,
whereas $\mathcal{P}$ simply represents the probability of generating that
partial ADT value.% (this probability can be easily calculated based on the
% generation frequencies used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewritten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P}^T)^{n}
\end{align*}
%
% The vector $\beta$ can be obtained by analyzing the shape (namely holes) of each
% source of structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as shown in Fig. \ref{fig:vectors}.
%
We note |simplify1| and |simplify2| to the patterns occurring in the first and
second clauses of |simplify|, respectively.
%(Dashed lines indicate the composition of the diffent sub-vectors
%that are calculated separately).

@@

Note that by varying the shape of the vector $\mathcal{P}$ we can tune the
distribution of our random generator in a way that can be always characterized
and predicted.
%
\dragenp follows a similar approach as \dragen and uses an heuristic to tune the
generation probabilities of each source of structural information.
%
This is done by running a simulation-based optimization process at compile-time.
%
This process is parameterized by the desired distribution of values set by the
user.
%
In this manner, developers can specify, for instance, a uniform distribution of
data constructors, pattern matching values and function calls or, alternatively,
a distribution of values with some constructions appearing in a different
proportion as others, e.g., two times more functions calls to |div| than |Join|
constructors.
